Investigating social change during cultural contact period using geometric morphometry of pottery shapes from Iron Age northeastern Taiwan

### 2020-01-15 analysis/paper/paper.Rmd
fix some problems about code and edit text, make the word count right

To identify craft specialization, the variation of standardization of specialized product is commonly used as indicator based on the assumption that specialized mass production will lead to uniformity of the product due to routinization, increased skills, and fewer number of producers involved in production [@Stark1995; @Costin1991; @Costin2001]. For ceramic standardization, several measurements have been proposed, such as metric, compositional, and technological approaches [@Arnold2000; @Boness2015; @Rice1991; @Blackman1993; @Costin1991; @Tite1999; @Roux2015]. Among those variables, metric measurement is widely applied in archaeological assemblages by calculating coefficient of variation for sets of data coupled with statistical comparison to identify lower value as an indicator of higher degree of standardization [@Eerkens2001; @Roux2003; @Stark1995]. @Roux2018's ethnographic study in the Jodhpur region of India suggests the number of artisans can be assessed by comparing coefficient of variation and the differences between artisans can be detected at both intra-individual level and inter-individual level in a region. Although this traditional typological and linear measurements are useful for identifying differences in archaeological assembles, it might be limited because dimensional measurements can be insensitive to subtle variations resulting from changes in assemblages. Shape analysis based on geometric morphometrics method are new approaches for measuring the degree of standardization since it takes the overall shape as variable for comparison [@Slice2007]. Using geometric morphometrics approaches, we study shape and standardization of locally made ceramics at Kiwulan (1350-1950 AD) [@Chen2007], a large multi-component archaeological site in northeastern Taiwan, to identify changes in ceramic production in cross-cultural interaction contexts. The aim of this paper is to investigate if there are increasing ceramic specialization resulting from interaction with the Europeans in the 17th century or the Chinese in the 19th century, two major foreign influences in early historical Taiwan, that might indicate social changes in the indigenous society.

@@

Outline approaches especially EFA is suitable for shapes lacking representative landmarks or curves convey more meaningful variations. The applications in archaeology are seen on human remains or zooarchaeology to understand the biological variation and recently more on stone artifacts to explore stone tool technology [@Fox2015; @Hoggard2019; @Iovictua2010]. Ceramics is another archaeological materials that outline approaches can provide new insights into shape variation to answer anthropological questions related to production. @Topi2018 examine 89 photographs of globular jars from the Casas Grandes of northwest Mexico using semi-landmarks method to explore the relationship between social complexity and the emergence of craft specialization. Their results suggest that some ceramic types from Medio period (AD 1200-1450), were made by specialists. Among those high standardized jars, some were made by attached specialists, while others were made by independent specialists according to distribution patterns of jars. @Wilczek2014 evaluate the concordance between outline-based approaches, EFA and Discrete Cosine Transform (DCT), and traditional typology by studying 154 complete ceramic vessels with varied shapes from the Bibracte oppidum in France. The results show the variation indicated by EFA and DCT matches traditional ceramic typology, which supports that outline-based approaches can be efficiently used for studying variations in ceramic shapes. Furthermore, the EFA results help to understand the level of production standardization over time across the region. Those examples show that outline approaches can distinguish variation in ceramic shapes in a finer resolution to understand deeply the issue of craft specialisation.

Taking the ceramics data from Kiwulan, northeaster Taiwan, we use EFA to evaluate the level of standardization of ceramics in relation to the European presence in the 17th century that might hint the emergence of ceramic specialization. In addition, we use significance test for the equality of coefficient of variation of shape variables to compare the vessel shapes from different periods in a statistical way. Using pottery shape as a proxy to study craft specialization, we answer the questions: Did foreign influence have impacts on indigenous pottery production that can be detected in the shape of the vessels? How does pottery change in shape in relation to foreign influences? Did pottery become more homogeneous and standardized in shape after foreign contacts with the European colonizer or the Chinese immigrants? Starting from the historical background of colonial influences and its reflection in the material record, this study explore the relationship between the emergence of social inequality and craft specialization that can extend our understanding of the reactions of indigenous societies in the periphery of colonial control, providing a perspective that could be applied in other parts of the world in a similar context.

@@

Ceramics analyzed in this paper come from 40 units (4m by 4m) sampled from four connected trenches at Kiwulan, northeastern Taiwan. Kiwulan is situated on a hill near a riverside at the northern margin of Yilan, which is characterized by a triangular alluvial plain facing eastwards the Pacific and mountains on three other sides. It is an ideal context to study peripheral colonial influence because it was isolated from intensive direct contact with the Spanish and Dutch colonial presence in other parts of Taiwan. The chronology of Kiwulan is divided into two cultural components, the upper and the lower component, with a sterile layer in between [@Chen2007]. This paper focuses on the upper component dating from AD 1350 to AD 1950, which covers the late Iron Age and the historical period defined by the European presence in Taiwan in the early 17th century. The Dutch firstly came and occupied southern Taiwan in 1624 and then the Spanish occupied northern Taiwan in 1626. In 1642, the Spanish was expelled by the Dutch, who took over their forts in northern Taiwan. Since then, western Taiwan was mostly under the colonial rule of the Dutch until 1662 when the Kingdom of Tungning in Taiwan was founded by Koxinga, a loyalist of the Ming dynasty of China [@Andrade2007].

The upper component experienced foreign contacts including the European colonial presence in the 17th century and great waves of Chinese immigrants since the 19th century. Imported ceramics from mainland China, stonewares, and ornaments elements such as beads were trade goods commonly found in the upper component that indicates the frequent long-distance trade activities in the 17th century during the European presence. In addition to artifacts, features such as burials, middens, and post-holes with in-situ posts that explain Kiwulan was a large settlement site [@Chen2007]. To compare different foreign influences, the deposits were classified into three phases, the pre-European, European, and Chinese periods. The principle of identification of three phases is based on 32 radiocarbon ages, excavation depth, consistency of contexts, and types of chronologically diagnostic artifacts, such as blue and white porcelains, light grey glazed jars, and large dark brown glazed stoneware jars prevailing in the 17th century, and bricks and tiles used by the Chinese in the 19th century [@Hsieh2009; @Wang2011]. The deposit shows signs of continuous human occupation during each of the three phases across the sampling area.

The predominant artefact found is locally made ceramics, which distributed throughout the sequence across the site. More than 550 thousands of sherds were excavated from the upper component, and around 1,200 vessels were reconstructed. It is interesting to note that there are only two forms of local vessels, one is cooking pot and another one is steamer made of two cooking pots stacked together with a filter clay layer in between. Those pots show high consistency in shape. They are globular body with short neck and wide mouth. The exterior surface below the neck is treated with a wide variety of impressed geometric motifs. They were believed used for cooking because of the evidence of charred residues and carbon deposits commonly observed on the interior and soot on the exterior. Moreover, a lack of evidence of other utilitarian earthenware ceramics found at this site suggests the globular pot was mainly used for daily cooking. They are fired to orange to brownish color with a fully oxidized core or a reduced core with oxidized fringes. Finger impressions and seams usually on the interior indicate they were pinched using slabs of clay and shaped by hand.

```{r metric-CVs-test}
library(here)
metric_cvs_with_phase <- readr::read_csv(here("analysis", "data", "raw_data", "Kwl_p_metrics_cv_test.csv"))

@@

The metric attributes for 362 reconstructed pots (pre-European = 153, European = 173, Chinese = 36) from the 40 sampling units were measure and compared using coefficient of variation (CV) with a significance test for the equality of CVs. The attributes include thickness of rim, neck, and body, and diameter of rim, neck, and body. For each pot, two measurements for thickness and diameter were took to compute an average value. To understand more about the pottery shape, we also compare the ratio of Rim/Body thickness, and the ratio of Rim/Body diameter. The results (Table \@ref(tab:metric-CVs-test))show that there is a significant difference (P-value < 0.01) in the variation of two metric attributes, the rim diameter and the neck diameter across three phases. The CV values show that the variation of rim diameter is small (6.8%) before European contact, increases after the European presence (9.6%), and decreases after the Chinese presence (6.9%). For the neck diameter, the CV values indicate smaller variation after the European period (7.7%) and the Chinese period (7.1%) compared to pre-European period(10.6%).

@@

A total of `r total_no` vessels from pre-contact contexts (n = 32), post-European contexts (n = 27), and Chinese contact contexts (n = 14) were selected for this study because of their completeness mostly covering parts from rim to bottom.

@@

@@

The elliptic Fourier coefficients of `r total_no` pottery from three phases were calculated. Reliable pottery outline was captured by 13 harmonics that gather 99 % of the total harmonic power. The average shape of vessels from each phase was visualized with the mean of the standardized Fourier coefficients within each phase group (Figure \@ref(fig:EFA-analysis)).

@@

```{r thin-plate-splines}
# Thin plate spline analysis (TPS)
# Thin plate splines (TPS) was used to compare average shapes and visualise the outline deformations required to pass from an extreme of the morphospace to the other.
# Deformation grids

@@

#? How to plot them next to each other, cowplot doesn't work when the plot not ggplot objects?
# plot_grid(a, b, labels = c('A', 'B'))

@@

```{r contribution-PCs, fig.cap= "Pottery shape variation based on scores along principal component axes. The rows from the top to the bottom represent the first to third principal component respectively. Each column shows the reconstructed shape variable according to the score mean ±2 standard deviation (SD) with a 0.5 interval"}
# the contributions of each PC to the shapes
pc_contrib <- PCcontrib(op.p, nax = 1:3) 
#?How to hide above plot, tried dev.off but seems it only stop to compute the following one
pc_contrib$gg +  
  geom_polygon(fill="slategrey", col="black") + 
  theme_minimal(base_size = 6)

@@

```{r first-two-PCs-byplot, fig.cap= "Pottery shape distribution by each occupation phase according to the first two PCs"}
# the pots groups by phase, and their shape distribution according to 
# the first two principle components of the EFA 
png(here::here("analysis/figures/pca-biplot.png"), res = 300, w = 1500, h = 1500)
plot(op.p, 2, 
     morpho = TRUE, 
     ellipses = TRUE,
     cex = 1, 
     cex.labelsgroups = 1.5)
dev.off()

@@

manv <- MANOVA_PW(op.p, "phase") #? how to hide the process, echo = FALSE?
su_manv <- as.data.frame(manv$summary)
# how to reduce the number in the table to only 4 digits, tried mutate with signif but doesn't work?
#su_manv %>% 
  #mutate('Pr(>F)' = signif('Pr(>F)', 4))

@@

```{r variance-test-PCs}
# we don't use CV because that is not good for measurements with no absolute zero
pc_plot %>% 
  group_by(pcn, phase) %>% 
  summarise(variance = var(pc))

@@

```{r CV-test-PCs_plot, fig.cap= "P-values of the CV equality test of PC1 and PC2 between phases, red color indicates p-value is less than 0.05"}

PC1_PC2_cv_tests %>%
  mutate(colour = if_else(p_value < 0.05, "red", "black")) %>% 
  ggplot(aes(str_wrap(phases, 40),
             p_value,
             colour = I(colour),
             shape = PC)) +
  geom_point(size = 8,  position = position_dodge(width = 0.5)) +
  theme_minimal(base_size = 12) +
  coord_cartesian(clip = 'off') +
  xlab("") +
  ylab("CV equality test p-values")

@@

The variation of pottery shape for each phase is explored by the distributions of the first three PC scores to compare the extent of standardization across phases. Figure \@ref(fig:PCs-distribution) suggests variations of PC scores for each vessel by phase. The first PC, the hight of the vessels, shows more variations in the pre-European period compared to the Chinese period. That indicates standardization appears strong in the Chinese period but weak in the pre-European period. The second PC also presents a similar pattern that stronger standardization of shape in the Chinese period compared to the other two phases. To see whether the differences in the distribution of PCs between any two phases is statistically significant, coefficients of variation statistic (CVs) among phase groups of PCs is examined with a modified signed-likelihood ratio test for the equality of CVs. P-values for the significant test of CVs of PC1 and PC2 show significant differences (P-value < 0.01) in the standardization of vessel shape across some periods, especially between Chinese contact with either pre-European or post-European(Figure \@ref(fig:CV-test-PCs_plot), Table \@ref(tab:CV-test-PCs-table)). Significant difference is also detected between pre-European and post-European (P-value < 0.05). Given the decline of the indigenous population in the Chinese period might be a cause of a more standardized shape [@Chen2007], we examined the spatial distribution of pottery selected in this paper. The pottery distributed across the sampling area without any clusters could rule out the bias caused by the population dynamics.

<---1. cv test for elliptic fourier coefficients 
2. PCA - Anova - statistical tests - common strategies for testing PCA
3. specific points, the outliers mean -->

@@

Previous investigations at Kiwulan suggest an unequal distribution pattern of prestige goods, trade ornaments specifically, after the presence of the European that hints the emergence of social inequality. We examine the degree of the ceramic standardization to measure ceramic specialization that is usually associate with social differentiation, such as the elite control [@Costin2001; @Junker1999]. The results confirm that the differences in pottery shapes can be detected using EFA method, indicating the foreign influences on the indigenous society- the Europeans in the 17th century and the Chinese in the 19th century. The result of MANOVA shows there is significant difference in shapes between pre-European and Post-European period, and between the pre-European and the Chinese period. The average shape presents round body with wider rim and neck before the European contact, and turns to oval-shaped body with narrower rim and neck after the European presence and even more obvious during the Chinese contact period. In general, pottery shapes become shorter in hight over time that leads to oval-shaped body. For the standardization of pottery shape, the result of CV test indicates the significant differences between pre-European period and post-European period (P-value < 0.05), and between Chinese contact and either pre-European or post-European (P-value < 0.01). This suggests pottery shape becomes more homogeneous and standardized after foreign contacts with the European colonizer and even more with the Chinese immigrants. The compositional analysis shows clay pastes are similar, regardless of the increasing standardization of the pottery shape, reflecting the raw material sources remains unchanged. More standardized shape in the Chinese period might indicate the appearance of specialized group. To figure out whether this change is related to the function of pot, we made an attempt to conduct lipid analysis for potsherds samples to identify the source of residue absorbed into the fabric. However, we did not obtain useful results due to a low lipid yield, which might result from the thin and less porous fabric that offers limited spaces to protect organic molecules from microbiological degradation [@Evershed2008, pp. 909].

@@

More standardized pottery shape without controlling the clay paste or technology suggests the shape as a preferred characteristic in pottery production during the Chinese period. However, the standardization is not reflected by most of the metric attributes, such as the rim, neck, and body thickness. There is higher variation in thickness attributes compared to diameter attributes, indicating shape is the main attribute that presents homogeneity. In addition to the emergence of specialized groups indicative of changes in social organization, a further explanation could be that producing homogeneous pottery is an act of resistance to the Europeans intrusion and the the Chinese immigrants by emphasizing the practice of traditional pottery production. In a culture contact situation, social identity could be expressed through materials practices to show cultural homogeneous and distinction from other groups [@Voss2005]. It is also important to recognize that social identity might be more complicated in a colonial context, not just a colonized–colonizer or local/foreign dichotomy [@Voss2005; @Voss2008]. The case study of Shamaoshan cemetery (3BC- 4AD) in Southwest China showing the process of incorporation of Southeast China into Han Empire involved a century of conflicts, resistance, and acceptance among social groups with different identity, especially in the context of Han immigrants [@Wu2019]. At Kiwulan, among the pottery attributes, shape is the most obvious and visible trait, which could be viewed as an expression of social boundary or identity [cf. @Roux2015].

- cultural resistance
- trade [@Alizadeh2018]
- identity, social boundary, visible trait, Raux2015, Boness2017
- thickness and diameter show difference but not reach the threshold of specialization, also the clay paste does not show any specific change throughout the time
- shape as an symbol

>>>>

To identify craft specialization, the variation of standardization of specialized product is commonly used as indicator based on the assumption that specialized mass production will lead to uniformity of the product due to routinization, increased skills, and fewer number of producers involved in production [@Stark1995; @Costin1991; @Costin2001]. For ceramic standardization, several measurements have been proposed and discussed, such as metric, compositional, and technological approaches [@Arnold2000; @Boness2015; @Rice1991; @Blackman1993; @Costin1991; @Tite1999; @Roux2015]. Among those variables, metric measurement is widely applied in archaeological assemblages by calculating coefficient of variation for sets of data coupled with statistical comparison to identify lower value as an indicator of higher degree of standardization [@Eerkens2001; @Roux2003; @Stark1995]. @Roux2018's ethnographic study in the Jodhpur region of India suggests the number of artisans can be assessed by comparing coefficient of variation and the differences between artisans can be detected at both intra-individual level and inter-individual level in a region. Although this traditional typological and linear measurements are useful for identifying differences in archaeological assembles, it might be limited because dimensional measurements can be insensitive to subtle variations resulting from changes in assemblages. Shape analysis based on geometric morphometrics method are new approaches for measuring the degree of standardization since it takes the overall shape as variable for comparison [@Slice2007]. Using geometric morphometrics approaches, we study shape and standardization of locally made ceramics at Kiwulan (1350-1950 AD) [@Chen2007], a large multi-component archaeological site in northeastern Taiwan, to identify changes in ceramic production in cross-cultural interaction contexts. The aim of this paper is to investigate if there are increasing ceramic specialization resulting from interaction with the Europeans in the 17th century or the Chinese in the 19th century, two major foreign influences in early historical Taiwan, that might indicate social changes in the indigenous society.

@@

Outline approaches especially EFA is suitable for shapes lacking representative landmarks or curves convey more meaningful variations. The applications in archaeology are seen on human remains or zooarchaeology to understand the biological variation and recently more on stone artifacts to explore stone tool technology [@Fox2015; @Hoggard2019; @Iovictua2010]. Ceramics is another archaeological materials that outline approaches can provide new insights into shape variation to answer anthropological questions related to production. @Topi2018 examine 89 photographs of globular jars from the Casas Grandes of northwest Mexico using semi-landmarks method to explore the relationship between social complexity and the emergence of craft specialization. Their results suggest that some ceramic types from Medio period (AD 1200-1450), were made by specialists. Among those high standardized jars, some were made by attached specialists, while others were made by independent specialists according to distribution patterns of jars. @Wilczek2014 evaluate the concordance between outline-based approaches, EFA and Discrete Cosine Transform (DCT), and traditional typology by studying 154 complete ceramic vessels with varied shapes from the Bibracte oppidum in France. The results show the variation indicated by EFA and DCT matches traditional ceramic typology, which supports that outline-based approaches can be efficiently used for studying variations in ceramic shapes. Furthermore, the EFA results help to understand the level of production standardization over time across the region. Those examples show that outline approaches can distinguish variation in ceramic shapes in a finer resolution to understand deeply the ceramic production.

Taking the ceramics data from Kiwulan, northeaster Taiwan, we use EFA to evaluate the level of standardization of ceramics in relation to the European presence in the 17th century that might hint the emergence of ceramic specialization. In addition, we use significance test for the equality of coefficient of variation of shape variables to compare the vessel standarization from different periods in a statistical way. Using pottery shape as a proxy to study craft specialization, we answer the questions: Did foreign influence have impacts on indigenous pottery production that can be detected in the shape of the vessels? How does pottery change in shape in relation to foreign influences? Did pottery become more homogeneous and standardized in shape after foreign contacts with the European colonizer or the Chinese immigrants? Starting from the historical background of colonial influences and its reflection in the material record, this study explore the relationship between the emergence of social inequality and craft specialization that can extend our understanding of the reactions of indigenous societies in the periphery of colonial control, providing a perspective that could be applied in other parts of the world in a similar context.

@@

Ceramics analyzed in this paper come from 40 units (4m by 4m) sampled from four connected trenches at Kiwulan, northeastern Taiwan. Kiwulan is situated on a hill near a riverside at the northern margin of Yilan, which is characterized by a triangular alluvial plain facing eastwards the Pacific with high mountains on three other sides. Yilan is an ideal context to study peripheral colonial influence because it was relatively isolated by natural barrier that reduced frequency of direct contact with the Europeans presences, the Spanish and the Dutch settled in northern Taiwan. The chronology of Kiwulan consists of two cultural components, the upper component and the lower component, with a sterile layer in between [@Chen2007]. This paper focuses on the upper component dated from AD 1350 to AD 1950, covering the late Iron Age and the historical period defined by the European presence in Taiwan in the early 17th century. The Dutch firstly came and occupied southern Taiwan in 1624 and then the Spanish occupied northern Taiwan in 1626. In 1642, the Spanish was expelled by the Dutch, who took over their forts in northern Taiwan. Since then, western Taiwan was mostly under the colonial rule of the Dutch until 1662 when the Kingdom of Tungning in Taiwan was founded by Koxinga, a loyalist of the Ming dynasty of China [@Andrade2007].

The upper component experienced foreign contacts including the Europeans in the 17th century and great waves of Chinese immigrants in the 19th century. Imported ceramics from mainland China, stonewares, and ornaments elements such as beads were trade goods commonly found in the upper component that indicates the frequent long-distance trade activities in the 17th century during the European presence in northern Taiwan. In addition to artifacts, features such as burials, middens, and post-holes with in-situ posts widespread across 1-2 m thick deposit explain that Kiwulan was a continuously occupied and large settlement site [@Chen2007]. To compare different foreign influences, the upper component were classified into three phases, the pre-European, European, and Chinese periods. The principle of identification of three phases is based on 32 radiocarbon ages, excavation depth, consistency of contexts, and types of chronologically diagnostic artifacts, such as blue and white porcelains, light grey glazed jars, and large dark brown glazed stoneware jars prevailing in the 17th century, and bricks and tiles used by the Chinese in the 19th century [@Hsieh2009; @Wang2011]. The deposit shows signs of continuous human occupation during each of the three phases across the sampling area.

The predominant artefact found is locally made ceramics, which distributed throughout the sequence across the site. More than 550 thousands of sherds were excavated, and around 1,200 vessels were reconstructed. It is interesting to note that there are only two forms of local vessels, one is cooking pot and another one is steamer made of two cooking pots stacked together with a filter clay layer in between. Those pots show high consistency in shape. They are globular body with short neck and wide mouth. The exterior surface below the neck is treated with a wide variety of impressed geometric motifs. They were believed used for cooking based on the evidence of charred residues and carbon deposits commonly observed on the interior and soot on the exterior. Moreover, a lack of evidence of other utilitarian earthenware ceramics suggests this globular pot was mainly used for daily cooking. They are fired to orange to brownish color with a fully oxidized core or a reduced core with oxidized fringes. Finger impressions and seams usually on the interior indicate they were pinched using slabs of clay and shaped by hand. In addition to Kiwulan site, this kind of pot was widely found at other archaeological sites during the same period in Yilan Plain.

```{r metricCVstest}
library(here)
metric_cvs_with_phase <- readr::read_csv(here("analysis", "data", "raw_data", "Kwl_p_metrics_cv_test.csv"))

@@

The metric attributes for 362 reconstructed pots (pre-European = 153, European = 173, Chinese = 36) from the 40 sampling units were measure and compared using coefficient of variation (CV) with a significance test for the equality of CVs. The attributes include thickness of rim, neck, and body, and diameter of rim, neck, and body. For each pot, two measurements for thickness and diameter were took to compute an average value. To understand more about the pottery shape, we also compare the ratio of Rim/Body thickness, and the ratio of Rim/Body diameter. The results (Table \@ref(tab:metricCVstest)) show that there is a significant difference (p-value less than 0.01) in the variation of two metric attributes, the rim diameter and the neck diameter across three phases. The CV values show that the variation of rim diameter is small (6.8%) before European contact, increases after the European presence (9.6%), and decreases after the Chinese presence (6.9%). For the neck diameter, the CV values indicate smaller variation after the European period (7.7%) and the Chinese period (7.1%) compared to pre-European period (10.6%).

@@

@@

A total of `r total_no` vessels from pre-contact contexts (n = 32), post-European contexts (n = 27), and Chinese contact contexts (n = 14) were selected for this study due to their completeness covering most parts from rim to bottom.

@@

The elliptic Fourier coefficients of `r total_no` pottery from three phases were calculated. Reliable pottery outline was captured by 13 harmonics that gather 99 % of the total harmonic power. The average shape of vessels from each phase was visualized by computing the mean of the standardized Fourier coefficients within each phase group (Figure \@ref(fig:EFA-analysis)). Figure \@ref(fig:thin-plate-splines) shows the shape changes described using thin-plate spline warping for paired periods, pre and post European periods, and post-European and Chinese period in a time sequence.

@@

```{r thin-plate-splines, fig.cap= "Compare the average shapes across phases using thin plate splines (TPS) with outline deformations required to pass from an extreme of the morphospace to the other."}
# Thin plate spline analysis (TPS)
# visualize the outline deformations required to pass from an extreme of the morphospace to the other
# Deformation grids

par(mfrow=c(1,2)) # plot them side by side

@@

@@

```{r contribution-PC-hide, results="hide", include= FALSE}
# the contributions of each PC to the shapes
pc_contrib <-  PCcontrib(op.p, nax = 1:3) 
```

```{r contribution-PCs, fig.cap= "Pottery shape variation based on scores along principal component axes. The rows from the top to the bottom represent the first to third principal component respectively. Each column shows the reconstructed shape variable according to the score mean ±2 standard deviation (SD) with a 0.5 interval"}
pc_contrib$gg +  
  geom_polygon(fill="slategrey", col="black") + 
  theme_minimal(base_size = 6)

@@

```{r first-two-PCs-byplot, fig.cap= "Pottery shape distribution by each occupation phase according to the first two PCs"}
# the pots groups by phase, and their shape distribution according to 
# the first two principle components of the EFA 
# png(here::here("analysis/figures/pca-biplot.png"), res = 300, w = 1500, h = 1500)
# plot(op.p, 2, 
#      morpho = TRUE, 
#      ellipses = TRUE,
#      cex = 1, 
#      cex.labelsgroups = 1.5)
# dev.off()

@@

manv <- rrtools:::quietly ( MANOVA_PW(op.p, "phase") ) #? how to hide the process, echo = FALSE?
su_manv <- as.data.frame(manv$summary)
# how to reduce the number in the table to only 4 digits, tried mutate with signif but doesn't work?
su_manv <- 
  su_manv %>% 
  mutate_all(round, 4)

@@

```{r variance-test-PCs}
# we don't use CV because that is not good for measurements with no absolute zero
xx <- 
  pc_plot %>% 
  group_by(pcn, phase) %>% 
  summarise(variance = var(pc))

@@

```{r CV-test-PCs-plot, fig.cap= "P-values of the CV equality test of PC1 and PC2 between phases, red color indicates p-value is less than 0.05"}

add_line_format <- function(x, ...){
  str_replace(x, " vs", "\nvs")
}

PC1_PC2_cv_tests %>%
  mutate(colour = if_else(p_value < 0.05, "red", "black")) %>% 
  ggplot(aes(str_wrap(phases, 40),
             p_value,
             colour = I(colour),
             shape = PC)) +
  geom_point(size = 8,  position = position_dodge(width = 0.5)) +
  theme_minimal(base_size = 12) +
  scale_x_discrete(labels = add_line_format(unique(sort(PC1_PC2_cv_tests$phases)))) +
  coord_cartesian(clip = 'off') +
  xlab("") +
  ylab("CV equality test p-values")

@@

The variation of pottery shape for each phase is explored by the distributions of the first three PC scores to compare the extent of standardization across phases. Figure \@ref(fig:PCs-distribution) suggests variations of PC scores for each vessel by phase. The first PC, the hight of the vessels, shows more variations in the pre-European period compared to the Chinese period. That indicates standardization appears strong in the Chinese period but weak in the pre-European period. The second PC also presents a similar pattern that stronger standardization of shape in the Chinese period compared to the other two phases. To see whether the differences in the distribution of PCs between any two phases is statistically significant, coefficients of variation statistic (CVs) among phase groups of PCs is examined with a modified signed-likelihood ratio test for the equality of CVs. P-values for the significant test of CVs of PC1 and PC2 show significant differences (P-value less than 0.01) in the standardization of vessel shape across some periods, especially between Chinese contact with either pre-European or post-European(Figure \@ref(fig:CV-test-PCs-plot), Table \@ref(tab:CV-test-PCs-table)). Significant difference is also detected between pre-European and post-European (P-value less than 0.05). Given the decline of the indigenous population in the Chinese period might be a cause of a more standardized shape [@Chen2007], we examined the spatial distribution of pottery selected in this paper. The pottery distributed across the sampling area without any clusters could rule out the bias caused by the population dynamics.

@@

Previous investigations at Kiwulan suggest an unequal distribution pattern of prestige goods, trade ornaments specifically, after the presence of the European that hints the emergence of social inequality. We examine the degree of the ceramic standardization to measure ceramic specialization that is usually associate with social differentiation, such as the elite control [@Costin2001; @Junker1999]. The results confirm that the differences in pottery shapes can be detected using EFA method, indicating the foreign influences on the indigenous society- the Europeans in the 17th century and the Chinese in the 19th century. The result of MANOVA shows there is significant difference in shapes between pre-European and Post-European period, and between the pre-European and the Chinese period. The average shape presents round body with wider rim and neck before the European contact, and turns to oval-shaped body with narrower rim and neck after the European presence and even more obvious during the Chinese contact period. In general, pottery shapes become shorter in hight over time that leads to oval-shaped body. For the standardization of pottery shape, the result of CV test indicates the significant differences between pre-European period and post-European period (P-value less than 0.05), and between Chinese contact and either pre-European or post-European (P-value less than 0.01). This suggests pottery shape becomes more homogeneous and standardized after foreign contacts with the European colonizer and even more with the Chinese immigrants. The compositional analysis shows clay pastes are similar, regardless of the increasing standardization of the pottery shape, reflecting the raw material sources remains unchanged. More standardized shape in the Chinese period might indicate the appearance of specialized group. To figure out whether this change is related to the function of pot, we made an attempt to conduct lipid analysis for potsherds samples to identify the source of residue absorbed into the fabric. However, we did not obtain useful results due to a low lipid yield, which might result from the thin and less porous fabric that offers limited spaces to protect organic molecules from microbiological degradation [@Evershed2008, pp. 909].

@@

More standardized pottery shape without controlling the clay paste or technology suggests the shape as a preferred characteristic in pottery production during the Chinese period. However, the standardization is not reflected by most metric attributes, such as the rim, neck, and body thickness. There is higher variation in thickness attributes compared to diameter attributes, indicating shape is the main attribute that presents homogeneity. In addition to the emergence of specialized groups indicative of changes in social organization, a further explanation could be that producing homogeneous pottery is an act of resistance to the Europeans intrusion and the the Chinese immigrants by emphasizing the practice of traditional pottery production. In a culture contact situation, social identity could be expressed through materials practices to show cultural homogeneous and distinction from other groups [@Voss2005]. It is also important to recognize that social identity might be more complicated in a colonial context, not just a colonized–colonizer or local/foreign dichotomy [@Voss2005; @Voss2008]. The case study of Shamaoshan cemetery (3BC- 4AD) in Southwest China showing the process of incorporation of Southeast China into Han Empire involved a century of conflicts, resistance, and acceptance among social groups with different identity, especially in the context of Han immigrants [@Wu2019]. Among the pottery attributes, shape is the most obvious and visible trait, which could be viewed as an expression of social boundary or identity [cf. @Roux2015].

- trade [@Alizadeh2018]
- identity, social boundary, visible trait, Raux2015, Boness2017
- thickness and diameter show difference but not reach the threshold of specialization, also the clay paste does not show any specific change throughout the time
- shape as an symbol
